Shared Subspace Model

### 2016-05-28 paper/mscov.tex
First complete draft of paper

% From Hoff Although the covariance matrices corresponding to different
% populations are unlikely to be exactly equal they can still exhibit a
% high degree of similarity. For example, some pairs of variables may be
% positively correlated across most groups, whereas the correlation
% between other pairs may be consistently negative. In such cases much
% of the similarity across covari- ance matrices can be described by
% similarities in their principal axes, which are the axes that are
% defined by the eigenvectors of the covariance matrices. Estimating the
% degree of across-popu- lation eigenvector heterogeneity can be helpful
% for a variety of estimation tasks.

  We develop model-based methods for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p > n$ setting.  This is done by assuming a spiked covariance model
  for each group, and sharing information about the subspace spanned
  by the group-level eigenvectors.  We propose an EM algorithm for
  estimating the shared subspace and the covariance matrix of the data
  projected onto this subspace.  We also provide a Gibbs sampling
  algorithm for estimating the posterior uncertainty of group-level
  eigenvectors and eigenvalues.  Finally, we demonstrate the utility
  of our method in analyses exploring covariability in biological data.

@@

% \section{NOTES}

% \begin{itemize}
% \item For MSCov to work in no-pooeld case need $R < S$ (doens't work if $S=P$).
% \item Explore the effect on estimation when R=2 two eigenvalues converge/diverge.
% \item TODO: Experiments: 
% \begin{itemize}
% \item S=20, R=2, data generated under no-pooled model.
% \item Fit with R=2 to R=20, Fix S=20
% \item Fit with R=2, S=5 to 20
% \item Fix $1/(1-Y^2)$ bug for Y=1
% \item debeta_un = 0 for large sample sizes (underflows)
% \item Check that CP models yield same results using fitBayesianSpike and fit-subspace.
% \end{itemize}
% \end{itemize}

@@

Introduction to covariance estimation and spiked covariance model.

Suppose $\bl S$ has a possibly degenerate Wishart$(\Sigma,n$) 
distribution with density 
given by 
\begin{equation} 
p(\bl S | \Sigma, n) \propto_{\Sigma} l(\Sigma:\bl S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1} \bl S/2 ) ,  
\label{eqn:lik}
\end{equation}
where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.

% However, when the population size is large and comparable with the
% sample size, as is in many contemporary data, it is known that the
% sample covariance matrix is no longer a good approximation to the
% covari- ance matrix. Baik

Such a likelihood results from $\bl S$ being, for example, a residual 
sum of squares matrix from a multivariate regression analysis. In  this case,
$n$ is the number of independent observations minus the 
rank of the design matrix. 
The spiked principle 
components model (spiked PCA) introduced by \citet{Johnstone2001} assumes that 
\begin{equation} 
\Sigma = \sigma^2 (  \bl U  \Lambda  \bl U^T  + \bl I )
\label{eqn:spiked}
\end{equation}

where $\Lambda$ is an $r\times r$ diagonal matrix and 
$\bl U \in \mathcal V_{r,p}$, with 
%$r< ( n \wedge p)$. 
$r<p$. Thus, the spiked covariance model is a low rank factor model
with homoscedastic error \citep{}.  Such a formulation is appealing because it explicitly
partitions the covariance into a tractable low rank ``signal'' and homscedastic ``noise''.

For such a covariance matrix, we have
$\Sigma^{-1} = (\bl U \Lambda \bl U^T + \bl I )^{-1} = \bl I - \bl U
\Omega \bl U^T $
where $\Omega = \text{diag}(\omega_1,\ldots, \omega_r)$ with
$\omega_s = \lambda_j/(\lambda_s+1)$.  The likelihood (\ref{eqn:lik})
in terms of $(\sigma^2,\bl U, \Omega)$ is then
\begin{equation}
L(\sigma^2,\bl U , \Omega : \bl Y)  = 
(\sigma^2)^{-np/2} \etr(-\bl S/[2\sigma^2]) \times 
   \etr( \Omega \bl U^T [\bl S/(2\sigma^2)] \bl U ) \times  
 \left\{ \prod_{s=1}^r (1-\omega_s)^{n/2} \right \} .  
\label{eqn:rplik}
\end{equation}

Classical results for parametric models 
(e.g., \citet{schwartz_1965})  
imply that 
asymptotically in $n$ for fixed $p$, a Bayes estimator
will be consistent for a spiked  population covariance 
as long as the assumed number of spikes is greater than or equal to 
the true number. 
However, the situation is more difficult 
when both $n$ and $p$ are large.  
%Since a Bayes estimator for an invariant loss function 
%is orthogonally equivariant under this prior, 
%it can be expressed as 
%$\hat \Sigma(\bl S) = \bl E \bl D(\bl L) \bl E^T$, 
%where $\bl E$ and $\bl L$ are the matrices of eigenvectors and 
%eigenvalues of $\bl S$, and $\bl D$ denotes a diagonal matrix 
%that is a function of $\bl L$. 
%As a result 
Under the spiked covariance model, 
results of 
\citet{Baik2006, Paul2007} indicate that if 
$p/n \rightarrow \gamma >0$ as $n\rightarrow \infty$,   
the top $r$ eigenvalues of $\bl S/(n\sigma^2)$ will converge 
to upwardly  biased versions of $(\lambda_1+1,\ldots, \lambda_r+1)$, 
if each $\lambda_s $ is greater than $\sqrt{\gamma}$.   
This has led several authors to suggest estimating 
$\Sigma$ via shrinkage of the eigenvalues of the sample covariance 
matrix. In particular, 
in the setting that $\sigma^2$ is known, 
\citet{Donoho2013} propose  
setting all sample eigenvalues that fall below 
the value $(1+\sqrt{\gamma})^2$ to be one, and shrinking the larger 
eigenvalues in a way that depends on the particular 
loss function being used. 
These shrinkage functions are shown to be asymptotically 
optimal in the $p/n\rightarrow \gamma$ setting.

As demonstrated by Gavish et al \citep{Gavish2014}, even when the
true rank is large, if blah it is still asymptotically preferable to fit a lower rank
blah .  it is always better than hard thresholding at any other value, and is always better than ideal truncated singular value decomposition (TSVD)

Bayes estimators under the prior distribution described above 
can also be viewed as eigenvalue shrinkage estimators: 
Since a Bayes estimator for an invariant loss function 
is orthogonally equivariant under this prior, 
it can be expressed as 
$\hat \Sigma(\bl S) = \bl E  \hat \Sigma (\bl L) \bl E^T$, 
where $\bl E$ and $\bl L$ are the matrices of eigenvectors and 
eigenvalues of $\bl S$. Furthermore, it is straightforward to 
show that $\hat \Sigma (\bl L)$ must be a diagonal matrix, 
and so the Bayes estimator is obtained by applying the 
shrinkage function $\hat\Sigma$ to the sample eigenvalues $\bl L$. 
Understanding of this shrinkage function could in principle 
be pursued by investigation of the posterior distribution of the 
eigenvalue parameters $\Lambda$, or equivalently their  
transformed values $\Omega$.

However, in many modern applications, it is not enough to
independently estimate an individual covariance matrix.  Rather, a
comparative analysis of covariance patterns across multiple groups of
observations would be most informative.  For example, in case
control-studies, mean level effects may be small relative to subject
variability.  Differences in the covariance
matrices may lead to better insights about the differences between groups.
Alternatively, when mean level differences are detectable, better
estimates of the covariability of features across groups may lead to improved
understanding of the mechanisms underlying these apparent mean level
differences.

With this situation in mind, Flury (1984) developed estimation and
testing pro-cedures for the ‘common principal components’ model, in
which a set of covariance matrices
\citep{Flury1987} and \citep{Schott1991, Schott1999} considered
cases in which only certain columns or subspaces of U are shared across populations, and \citep{Boik2002} described a very general model in which eigenspaces can be shared between all or some of the populations.

Continuously parameterized covariance matrices. \citep{Hoff2011}, \citep{Chiu1996}, \citep{Yin2010}

In such models, it is no longer reasonable to assume orthogonal equivariance .
To address these issues, \citet{Hoff2009} introduced a model for
eigenvector shrinkage based on the matrix Bingham distribution.

Partial least squares discriminant analysis refs?
\citep{Westerhuis2008, Szymanska2012, Eriksson2001, Barker2003, Saebo2008}

Recently there has been significant interest gaussian graphical
models, .   In gaussian graphical models, zeros in the
precision matrix correspond to conditional independences between pairs
of features given the remaining features \citep{}.  \citet{Witten2014} developed a
method for pooled estimation inverse covariance matrices across
multiple groups.  Identifying a shared pattern of zero's in the inverse covariance
matrices for multiple classes of observations.

In the following section, we
propose a model for estimating and exploring differences between
covariance matrices from multiple groups of related data.

Below we describe our approach for comparaing covariance matrices on
a common low dimensional subspace.

@@

Assume we have $k=1, ..., K$ groups of matrices $\bl S_k$ each of
which follow a (degenerate) Wishart distribution, as in Equation
\ref{eqn:lik}.  Each group consists of of $n_k$ measurements of $p$
features, typically with $n_k << p$.  We pool information about the
covariance matrices by assuming that most of the relevant variation
for each group shares a low rank subspace spanned by
the columns of a matrix $V$.  That is,

\begin{equation}
\Sigma_k = V\psi_kV^T + \sigma^2_kI
\label{eqn:sspsi}
\end{equation}

Here, $V \in \mathcal{V}_{s,p}$ and $p \geq s$.  $V$ is only
identifiable up to right rotations.  Consequently,
$VV^T \in \mathcal{G}_{s, p}$, which defines the plane of variation
shared by all groups, is identifiable.  $\psi_k$ is a rank-$r$
covariance matrix on the $s$-dimensional subspace.  Such a model is
reasonable for many applications involving groups of highly structure
data on the same set of $p$ features.  For instance, in biological
applications, only certain subsets of molecules can physically
interact which constrains the space of possible correlation patterns.
Our model exploits this notion, by estimating differences between
groups on a common lower dimensional subpace.

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Ratio}
        \label{fig:dmelanRatio}
    \end{subfigure}
    \qquad     \qquad     \qquad 
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}r
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{Posterior}
        \label{fig:dmelanPosterior}
    \end{subfigure}
    \caption{ }
\end{figure}

\section{Inference}

In this section we outline Bayesian inference algorithms including
both an optimization algorithm and an MCMC algorithm for estimating
covariance parameters and their associated uncertainty.  MCMC
inference is particularly appealing because of the ease with which we
can quantify the uncertainty associated with each estimate.
Uncertainty quantification is particularly important in the $n << p$
setting where seemingly large differences in the point estimates of
the low rank covariance matrices across groups may not actually
reflect significant differences.  Although, uncertainty quantification
for all parameters is desirable, for the size of the problems we are
generally interested in, we found MCMC inference for the subspace
$VV^T$ to be intractable.  Our optimziation based approached is
signifcantly faster and works well for large large dimensional
datasets.  As such, we start, in Section \ref{sec:em}, by outlining an
expectation-maximization algorithm for estimating the shared subspace
and derving point estimates of the projected covariance $\psi_k$. In
Section \ref{sec:bayes} we describe a Gibbs sampler for posterior
inference of $\psi_k$ given the common subspace.  These two algorithms
can be used in conjuction to first identify and evaluate a shared
subspace and then to quantify the uncertainty associated with the
projected covariance matrices on this space.

\subsection{An Expectation-Maximization Algorithm}
\label{sec:em}

In this section we describe a computationally efficient algorithm for
estimating the shared subspace, $VV^T$, projected covariance matrix
$\psi_k$ and noise parameter $\sigma_k^2$.  First, note that the full
likelihood for the shared subspace model can be written as

\begin{align}
p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}\mathbf{S}_k/2)  \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\psi_kV^T +
  I))^{-1}\mathbf{S}_k/2) \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]\mathbf{S}_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]\mathbf{S}_k/2) 
\end{align}

Where we define $M_k^{-1} = (\psi_k + I) ^{-1}/\sigma_k^2$.  The log-likelihood in
$V$ (up to additive constant) is

@@

We propose an expectation-maximization (EM)
algorithm, to estimate the relevant quantities.  In particular we
maximize the marginal posterior distribution of $V$, treating 
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as the ``missing'' parameters to
be imputed.  We assume independent Jeffreys' prior distributions for
both $\frac{1}{\sigma_k^2}$ and $M_k^{-1}$ and a uniform prior for $V
\in \mathcal{V}_{s, p}$.  In particular we let
$p(\frac{1}{\sigma_k^2}) \propto \sigma_k^2$ and
$p(M_k^{-1}) \propto |M_k^{-1}|^{-(s+1)/2}$.  From the likelihood it
can easily be shown that the conditional posterior for $M_k^{-1}$ is

@@

\paragraph{Evaluating a Shared Subspace}

The above EM algorithm provides an estimate of a shared subspace, but
in practice there are no guarantees that the inferred subspace is
equally appropriate for all groups.  Moreover, the shared subspace can
be chosen in other ways, including more ad hoc methods
involving the singular value dcomposition or simply by incorporating
subject matter considerations.  Regardless of how a particular
subspace is identified, it is usually important to quantify how useful
this subspace is for explaining the variability in each group.  In
particular, if $\hat{V}$ is a basis for a subspace shared across
groups, then for each $k$, most of the variation in $\Sigma_k$ should
lie on the subspace spanned by $\hat{V}$.  To quantify the extent to
which this is true for different groups, we propose a simple estimator
for the proportion of ``signal'' variance that lies on any given
subspace.  Specifically, we apply the following estimator for ratio of
the sum of the first $s$ eigenvalues of
$\hat{V}^T\bl \Sigma_k \hat{V}$ to the sum of the first $s$
eigenvalues of $\Sigma_k$:

\begin{equation}
 \frac{||Y_k^T\hat{V}||_F}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{sup}
  ||Y_k^T\widetilde{V}||_F - \hat{\sigma}_k^2ps}
\label{eqn:ratio}
\end{equation}

>>>>

% From Hoff Although the covariance matrices corresponding to different
% populations are unlikely to be exactly equal they can still exhibit a
% high degree of similarity. For example, some pairs of variables may be
% positively correlated across most groups, whereas the correlation
% between other pairs may be consistently negative. In such cases much
% of the similarity across covariance matrices can be described by
% similarities in their principal axes, which are the axes that are
% defined by the eigenvectors of the covariance matrices. Estimating the
% degree of across-popu- lation eigenvector heterogeneity can be helpful
% for a variety of estimation tasks.

  We develop a model-based method for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p > n$ setting.  This is done by assuming a spiked covariance model
  for each group, and sharing information about the subspace spanned
  by the group-level eigenvectors.  We propose an EM algorithm for
  estimating the shared subspace and the covariance matrix of the data
  projected onto this subspace.  We also provide a Gibbs sampling
  algorithm for estimating the posterior uncertainty of group-level
  eigenvectors and eigenvalues.  Finally, we demonstrate the utility
  of our method in an analysis of biological data.

@@

@@

It has long been known that the sample covariance matrix is a poor
estimator of the population covariance matrix, particularly when the
sample size is comparable to or smaller than the number of features
($n \leq p$) \citep{Dempster1969, Stein1975}.  A host of estimators
have been devised to address this fact. These primarily include of
shrinkage based approaches \citep{Schafer2005, Ledoit2011} and low
rank factor models \citep{Mardia1980, Ullman2003, Fan2008}.  In many
cases, the covariance matrices are highly structured and thus can be
sparsely parameterized even if the matrix itself is not low rank
\citep{Burg1982, Williams1993} .  Along these lines, there has been a
recent interest in sparse Gaussian graphical model, in which the there
are assumed to be many zeros in the inverse covariance matrix
\citep{Meinshausen2006, Friedman2008}.

Most of these approaches assume that the goal is to estimate a
covariance matrix from a single dataset.  In many applications,
however, the data can be split into groups, each of which represent
samples from populations with distinct (but possibly related)
covariance matrices.  In this case, a comparative analysis of
covariance patterns across multiple groups of observations is often
warranted.  Evaluating similarities and differences between covariance
matrices can be an important complement to classical analyses which
center primarily on estimating differences in means across groups.
For instance, in case control-studies, mean level effects may be small
relative to subject variability; distributional differences between
groups may still be evident as differences in the covariances between
features.  Even when mean level differences are detectable, better
estimates of the covariability of features across groups may lead to
an improved understanding of the mechanisms underlying these apparent
mean level differences.  Further, accurate covariance estimation is an
essential part of many prediction tasks (e.g. quadratic discriminant
analysis).

By exploiting similarity across related groups of high-dimensional
data, we can increase the efficiency of existing covariance
estimators.  Some of the earliest approaches for modeling multiple
covariance matrices focused on estimation in terms of the spectral
decompositions for these matrices.  \cite{Flury1987} developed
estimation and testing procedures for the ``common principal
components'' model, in which a set of covariance matrices were assumed
to share the same eigenvectors.  \citet{Schott1991, Schott1999}
considered cases in which only certain eigenvectors are shared across
populations, and \citet{Boik2002} described an even more general model
in which eigenvectors can be shared between all or some of the groups.
More recently, \citet{Hoff2009}, noting that eigenvectors are unlikely
to be shared exactly between groups, introduced a hierarchical model
for eigenvector shrinkage based on the matrix Bingham distribution.
There has also been a significant interest in estimating covariance
matrices using Gaussian graphical models. For Gaussian graphical
models, zeros in the precision matrix correspond to conditional
independences between pairs of features given the remaining features
\citep{Meinshausen2006}.  \citet{Witten2014} extended existing work in
this area to the multi-group setting, by pooling information about the
pattern of zeros in across precision matrices.

% When the
% data can be grouped according to a set of continuous covariates
% continuously parameterized covariance models can be appropriate
% \citep{Chiu1996, Yin2010, Hoff2011}.

Another popular method for modeling relationships between
high-dimensional multivariate data related that is closely related to
low rank covariance estimation, is known partial least squares
regression (PLS). This approach, which is a special case of a bilinear
factor model, involves projecting the data into a lower dimension
space which maximizes the similarity of the two groups; this technique
does not require the data from each group to share the same feature
set.  An oft used variant for prediction, partial least squares discriminant analysis
(PLS-DA) is especially common in the chemometrics and bioinformatics
 \citep{Barker2003} .  Although closely related to
the approaches we will consider here, the primarily goal of of PLS based models
is to create regression or discrimination models, not to explicitly
infer covariance matrices from multiple groups of data.  Nevertheless,
the basic idea that data can often be well represented on a low
dimensional space is an appealing one that we leverage as well.

In many natural systems, high dimensional data is often very
structured and thus can be well understood on a relatively low
dimensional subspace. For example, with gene expression data, the
effective dimensionality is thought to scale with the number of gene
regulatory modules, not the number of genes themselves
\citep{Heimberg2016}.  As such, differences in gene expression across
groups should be expressed in terms of differences between these
regulatory modules rather than strict differences between expression
levels.  Such differences can be examined on a subspace that reflects
the correlations resulting from these modules.

Motivated by this notion, in this paper we propose a multi-group
covariance estimation model by sharing information about the subspace
spanned by group-level eigenvectors.  The shared subspace assumption
can be used to improve estimates and facilitate the interpretation of
differences between covariance matrices across groups.  For each
group, we assume ``the spiked covariance model'', a well studied
variant of the factor model \citep{Johnstone2001}.  In Section
\ref{sec:shared} we briefly review the behavior of spiked covariance
models for estimating a single covariance matrix before introducing
our extension to the multi-group setting.  In Section
\ref{sec:inference} we describe an optimization algorithm for
inferring the shared subspace and then outline a MCMC algorithm for
estimating the posterior distribution of the projected covariance
matrices.  Finally, in Section \ref{sec:simulation} we investigate the
behavior of these models in simulation before demonstrating the
utility of this model with real biological data in Section
\ref{sec:app}.

@@

\paragraph{The Spiked Covariance Model:}

Suppose $\bl S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(\bl S | \Sigma, n) \propto_{\Sigma} l(\Sigma:\bl S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1} \bl S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $\bl S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
introduced by \citet{Johnstone2001} assumes that
\begin{equation} 
\Sigma = \sigma^2 (  \bl U  \Lambda  \bl U^T  + \bl I )
\label{eqn:spiked}
\end{equation}
%
\noindent where $\Lambda$ is an $r\times r$ diagonal matrix and
$\bl U \in \mathcal V_{r,p}$, with $r \ll p$. Here $\mathcal V_{r,p}$,
known as the Stiefel manifold, comprises the set of $p \times r$
orthonormal matrices in $\mathcal{R}^p$.  The spiked covariance model
is thus a low rank factor model with homoscedastic error.  Such a
formulation is appealing because it explicitly partitions the
covariance matrix into a tractable low rank ``signal'' and isotropic
``noise''.

%  For such a covariance matrix, we have
% $\Sigma^{-1} = (\bl U \Lambda \bl U^T + \bl I )^{-1} = \bl I - \bl U
% \Omega \bl U^T $
% where $\Omega = \text{diag}(\omega_1,\ldots, \omega_r)$ with
% $\omega_s = \lambda_j/(\lambda_s+1)$.  The likelihood (\ref{eqn:lik})
% in terms of $(\sigma^2,\bl U, \Omega)$ is then
% \begin{equation}
% L(\sigma^2,\bl U , \Omega : \bl Y)  = 
% (\sigma^2)^{-np/2} \etr(-\bl S/[2\sigma^2]) \times 
%    \etr( \Omega \bl U^T [\bl S/(2\sigma^2)] \bl U ) \times  
%  \left\{ \prod_{s=1}^r (1-\omega_s)^{n/2} \right \} .  
% \label{eqn:rplik}
% \end{equation}

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, a Bayes estimator will
be consistent for a spiked population covariance as long as the
assumed number of spikes (non-unit eigenvalues) is greater than or
equal to the true number.  However, when $p$ is large relative to $n$,
as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
$p/n \rightarrow \gamma >0$ as $n\rightarrow \infty$, the top $r$
eigenvalues of $\bl S/(n\sigma^2)$ will converge to upwardly biased
versions of $(\lambda_1+1,\ldots, \lambda_r+1)$, if each $\lambda_s $
is greater than $\sqrt{\gamma}$ \citep{Baik2006, Paul2007} .  This has
led several authors to suggest estimating $\Sigma$ via shrinkage of
the eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose setting
all sample eigenvalues that fall below the value
$\sigma^2(1+\sqrt{\gamma})^2$ to be $\sigma^2$, and shrinking the
larger eigenvalues in a way that depends on the particular loss
function being used.  These shrinkage functions are shown to be
asymptotically optimal in the $p/n\rightarrow \gamma$ setting.

% As demonstrated by Gavish et al \citep{Gavish2014}, even when the
% true rank is large, if blah it is still asymptotically preferable to fit a lower rank
% blah .  it is always better than hard thresholding at any other value, and is always better than ideal truncated singular value decomposition (TSVD)

Many eigenvalue shrinkage estimators can be viewed in the Bayesian
context as derivative of a Bayes estimator under a particular prior.
Note that in the absence of real prior information, invariance
considerations suggest the use of priors on $\sigma^2$ and $\bl U$
that lead to equivariant estimators.  For instance, an (improper)
prior having density $p(\sigma^2,\bl U) = 1/\sigma^2 $, with respect
to the product of Lebesgue measure on $\mathbb R^+$ and the uniform
probability measure on $\mathcal V_{r,p}$, leads to Bayes estimators
$\hat \Sigma(\bl S)$ that are equivariant with respect to rotations
and scale changes.  For multi-group covariance estimation, however, we
often have reason to believe that a priori the eigenvectors are
related across groups.  In this case, group-level equivariance to
rotations is an unreasonable assumption.  Here, both eigenvalue
\emph{and} eigenvector shrinkage can play an important role in
improving covariance estimates.  Below, we introduce our extension of
the spiked covariance model to the group setting by including an
assumption about the space shared by the span of the group-level eigenvectors.

\paragraph{The Shared Subspace Model:}

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathcal{R}^3$}
        \label{fig:dmelanRatio}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $R^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}

The spiked covariance model assumes that the goal is to estimate a
single covariance matrix.  The situation is different when we are
interested in estimating multiple covariance matrices from distinct,
but related groups.  For multi-group estimation, we can often improve
upon the efficiency of covariance estimators by pooling information
across the groups of data.  Here, we assume we have $k=1, ..., K$
groups of mean-zero normal data, $Y_k$, where $Y_k$ consists of $n_k$
independent rows of measurements on $p$ features, typically with
$n_k \ll p$.  Then, $Y_k^TY_k = \bl S_k$ follow a (degenerate) Wishart
distribution as in Equation \ref{eqn:lik}.  To improve estimation, we
seek estimators for the population covariance from one group that depend on the
data from all $K$ groups, e.g. of the form
$\hat \Sigma_k(\bl S_1, ... \bl S_K)$.  To this end, we extend the
spiked covariance model to grouped data, by assuming that the
anisotropic variability from each group can be expressed on a common
low dimensional subspace.  Specifically, we posit that the covariance
matrix for each group can be written as

\begin{equation}
\Sigma_k = V\psi_kV^T + \sigma^2_kI
\label{eqn:sspsi}
\end{equation}
%
\noindent Here, the columns of $V \in \mathcal{V}_{p, s}$ determine a subspace
shared by all groups $k$ with $p \geq s$.  Throughout this paper we
will denote to the shared subspace as $VV^T \in \mathcal G_{r,p}$, where
$\mathcal G_{r,p}$, known as the Grassmann manifold, consists of all
$s$-dimensional linear subspaces of $\mathcal{R}^p$
\citep{Chikuse2012}.  Although $V$ is only identifiable up to right
rotations, $VV^T$, which defines the plane of variation shared by all
groups, is identifiable.  Later, to emphasize the connection to the
spiked PCA model (\ref{eqn:spiked}), we will write $\psi_k$ in terms
of its eigendecomposition, $\psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\psi_k$ (see
Section \ref{sec:bayes}).

Note that for the shared subspace model,
$V^T\Sigma_kV = \psi_k + \sigma_k^2I$ is an anisotropic
$s$-dimensional covariance matrix for the projected data, $Y_kV$.
In contrast, the data projected onto the orthogonal space,
$Y_kV_{\perp}$, is isotropic, and typically small, for all
groups.  In Figure \ref{fig:shared} we provide a simple illustration
of this fact using simulated $4$-dimensional data from two groups.  The
differences in distribution between the groups of data can be
expressed on a two dimensional subspace spanned by the columns of
$V \in \mathcal{V}_{4, 2}$.  Differences in the correlations between
the two groups manifest themselves on this shared subspace, whereas
only the magnitude of the isotropic variability can differ between groups
on the orthogonal space.  Thus, the the shared subspace model can be
viewed as a covariance partition model, where one partition includes
the anisotropic variability from all groups and the other partition is
constrained to the isotropic variability from each group.  This
isotropic variability is often taken to be ``measurement noise''.

%Such a modelis reasonable for many applications involving groups of
%highly structure data on the same set of $p$ features.  % For instance,
% in biological applications, only certain subsets of molecules can
% physically interact which constrains the space of possible correlation
% patterns.  Our model exploits this notion, by estimating differences
% between groups on a common lower dimensional subpace.

\section{Inference}
\label{sec:inference}
In this section we outline Bayesian inference algorithms including
both an optimization algorithm and an MCMC algorithm for estimating
covariance parameters and their associated uncertainty.  MCMC
inference is particularly appealing because of the ease with which we
can quantify the uncertainty associated with each estimate.
Uncertainty quantification is particularly important in the $n \ll p$
setting where seemingly large differences in the point estimates of
the low rank covariance matrices across groups may not actually
reflect significant differences.  Although uncertainty quantification
for all parameters is desirable, for the size of the problems we are
generally interested in, we found MCMC inference for the subspace
$VV^T$ to be intractable.  Our optimization based approached is
significantly faster and works well for high-dimensional datasets.  As
such, we start, in Section \ref{sec:em}, by outlining an
expectation-maximization algorithm for estimating the shared
subspace. In Section \ref{sec:bayes} we describe a Gibbs sampler for
posterior inference of $\psi_k$ given the common subspace.  These two
algorithms can be used in conjunction to first identify and evaluate a
shared subspace and then to quantify the uncertainty associated with
the projected covariance matrices.

\subsection{Estimating the Shared Subspace}
\label{sec:em}

In this section we describe a computationally efficient algorithm for
estimating the shared subspace, $VV^T$, the projected covariance matrix
$\psi_k$ and noise parameter $\sigma_k^2$.  First, note that the full
likelihood for the shared subspace model can be written as

\begin{align}
p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}\mathbf{S}_k/2)  \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\psi_kV^T +
  I))^{-1}\mathbf{S}_k/2) \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]\mathbf{S}_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]\mathbf{S}_k/2) 
\end{align}
%
\noindent where we define $M_k^{-1} = (\psi_k + I) ^{-1}/\sigma_k^2$.  The log-likelihood in
$V$ (up to additive constant) is

@@

We propose an expectation-maximization (EM) algorithm, to estimate the
relevant quantities.  In particular we maximize the marginal posterior
distribution of $V$, treating $M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as
the ``missing'' parameters to be imputed.  We assume independent
Jeffreys' prior distributions for both $\frac{1}{\sigma_k^2}$ and
$M_k^{-1}$ and a uniform prior for $V \in \mathcal{V}_{s, p}$.
Jeffreys' prior for these quantities corresponds to
$p(\frac{1}{\sigma_k^2}) \propto \sigma_k^2$ and
$p(M_k^{-1}) \propto |M_k^{-1}|^{-(s+1)/2}$.  From the likelihood it
can easily be shown that the conditional posterior for $M_k^{-1}$ is

@@

\paragraph{Evaluating a Shared Subspace:}

The above EM algorithm provides an estimate of a shared subspace, but
in practice there are no guarantees that the inferred subspace is
appropriate for all groups.  Moreover, the shared subspace can be
chosen in other ways, including more ad hoc methods involving the
singular value decomposition or simply by incorporating subject matter
considerations.  Regardless of how a particular subspace is specified,
it is important to quantify how useful this subspace is for explaining
the variability in each group.  In particular, if $V$ is a basis for
the shared subspace, then for each $k$, most of the variation in the
data should be preserved when projecting the data to this subspace.
To quantify the extent to which this is true for different groups, we
propose a simple estimator for the proportion of ``signal'' variance
that lies on any given subspace.  Specifically, for any inferred
subspace $\hat{V}$, we apply the following estimator for ratio of
  the sum of the first $s$ eigenvalues of
  $\hat{V}^T\bl \Sigma_k \hat{V}$ to the sum of the first $s$
  eigenvalues of $\Sigma_k$:
%
\begin{equation}
 \frac{||Y_k\hat{V}||_F}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{sup}
  ||Y_k\widetilde{V}||_F - \hat{\sigma}_k^2ps}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||.||_F$ is the Frobenius norm. If the inferred
subspace is in fact the true subspace, i.e. $\hat{V}\hat{V}^T = VV^T$,
then this estimator will asymptotically approach one.  To see this,
note that
$\underset{V \in \mathcal{V}_{p, s}}{sup}\frac{||Y_kV||_F}{n_k}$ is
equivalent to the sum of the first $s$ eigenvalues of the sample
covariance matrix $\frac{\bl S_k}{n_k}$.  With $\hat{\lambda}^{(k)}_i$
the $i$-th eigenvalue of $\frac{\bl S_k}{n_k}$, \citet{Baik2006}
demonstrated that asymptotically as $p, n_k \rightarrow \infty$ with
$\frac{p}{n_k} = \gamma_k$ fixed
